---
title: "3: Data Wrangling"
date: today
execute:
  eval: true
  message: false
  warning: false
---



## Executive Summary

This post cleans and analyzes messy sales and customer data, fixing inconsistent dates, currency formatting, and text casing. The analysis found that sales grew steadily from January to March, with Services as the top-performing category and Mid-Market customers generating the highest average sale per person, while also flagging one transaction with an invalid customer ID for follow-up.
---

## Introduction

As an accounting professional, you'll often receive data that needs cleaning before analysis. Common issues include:

- Inconsistent date formats (01/15/2027 vs 2027-01-15)
- Mixed number formats ($1,234.56 vs 1234.56)
- Inconsistent text (ACME CORP vs Acme Corp vs acme corp)
- Data stored in multiple systems that need to be combined

In this blog post, you will:

- Identify data quality issues
- Clean messy accounting data
- Combine data from multiple sources
- Create professional summaries


## Setup and Load Libraries

### Required Libraries

```{r}
#| label: setup
# Load required packages
library(tidyverse)  # For data manipulation
library(scales)     # For formatting numbers
```

### Creating Sample Data

Let's create sample data that shows common accounting data problems:

**Sales Transaction Data**

```{r}
#| label: sales-data
# Sales data with typical formatting issues
sales_raw <- tribble(
  ~transaction_id, ~customer_id, ~sale_date,   ~amount,      ~product_category,
  "TXN-001",       "CUST001",    "01/15/2027", "$11,250.00", "Software",
  "TXN-002",       "CUST002",    "01/20/2027", "19750",      "Services",
  "TXN-003",       "CUST003",    "2027-02-01", "$6,000.50",  "Software",
  "TXN-004",       "CUST001",    "02/15/2027", "33000.00",   "services",
  "TXN-005",       "CUST999",    "02/28/2027", "$4,750",     "Services",        # Bad customer
  "TXN-006",       "CUST004",    "3/10/2027",  "5000",       "Hardware",
  "TXN-007",       "CUST002",    "03/15/2027", "$14,000",    "Software",
  "TXN-008",       "CUST003",    "2027-03-20", "26250.00",   "HARDWARE"
)

# View the raw data
sales_raw
```

**Customer Master Data**


```{r}
#| label: customer-data
# Customer data with inconsistent formatting
customers_raw <- tribble(
  ~customer_id, ~customer_name,      ~region, ~segment,
  "CUST001",    "acme corporation",  "North", "Enterprise",
  "CUST002",    "GLOBEX INC",        "South", "Mid-Market",
  "CUST003",    "umbrella corp",     "West",  "",           # Missing segment
  "CUST004",    "Stark Industries",  "East",  "Enterprise",
  "CUST005",    "wayne enterprises", "North", "Enterprise"  # No sales in Q1
)

# View the raw customer data
customers_raw
```


### Identifying Data Quality Issues
Before cleaning, let's identify the problems:

**Check Date Formats**

```{r}
#| label: check-dates
# Look at the different date formats
sales_raw |>
  select(sale_date) |>
  distinct()
```

**Check Amount Formats**

```{r}
#| label: check-amounts
# Look at the different amount formats
sales_raw |>
  select(amount) |>
  distinct()
```

**Check Customer Name Consistency**

```{r}
#| label: check-names
# Look at customer name formatting
customers_raw |>
  select(customer_name)
```


**Find Missing Data**



```{r}
#| label: check-missing
# Check for blank segments
customers_raw |>
  filter(segment == "")
```

### Cleaning the Data

#### Step 1: Clean Sales Data

```{r}
#| label: clean-sales
# Clean the sales data
sales_clean <- sales_raw |>
  mutate(
    # Fix dates: parse_date_time can handle multiple formats
    sale_date = parse_date_time(sale_date, orders = c("mdy", "ymd")) |>
                as_date(),

    # Fix amounts: remove $ and , then convert to number
    amount = str_remove_all(amount, "[$,]") |>  # Remove $ and ,
             as.numeric(),

    # Standardize text to Title Case
    product_category = str_to_title(product_category),

    # Add useful columns for reporting
    month_name = month(sale_date, label = TRUE, abbr = FALSE)
  )
# View cleaned sales data
sales_clean
```


#### Step 2: Clean Customer Data

```{r}
#| label: clean-customers
# Clean the customer data
customers_clean <- customers_raw |>
  mutate(
    # Standardize customer names to Title Case
    customer_name = str_to_title(customer_name),

    # Replace blank segments with "Unclassified"
    segment = if_else(segment == "", "Unclassified", segment)
  )

# View cleaned customer data
customers_clean
```


### Combining the Data

#### Join Sales with Customer Information

```{r}
#| label: join-data
# Combine sales and customer data
# Use left_join to keep all sales, even if customer not found
sales_complete <- sales_clean |>
  left_join(customers_clean, by = "customer_id")

# View the combined data
glimpse(sales_complete)
```


#### Check for Data Quality Issues

```{r}
#| label: problem-sales
# Find sales without valid customer info
problem_sales <- sales_complete |>
  filter(is.na(customer_name)) |>  # Check for missing customer_name
  select(transaction_id, customer_id, amount, sale_date)

problem_sales

```


### Analysis and Insights

#### Monthly Sales Summary

```{r}
#| label: monthly-summary
# Calculate monthly sales (excluding problem transactions)
monthly_sales <- sales_complete |>
  filter(!is.na(customer_name)) |>  # Only valid customers
  group_by(month_name) |>  # Group by month_name
  summarise(
    total_sales = sum(amount),
    transaction_count = n(),
    average_sale = mean(amount)
  )

# View monthly summary
monthly_sales

# Format as currency
monthly_sales |>
  mutate(
    total_sales = dollar(total_sales),
    average_sale = dollar(average_sale)
  )

```

#### Sales by Product Category

```{r}
#| label: product-summary
# Analyze sales by product
product_summary <- sales_complete |>
  filter(!is.na(customer_name)) |>
  group_by(product_category) |>  # Group by product_category
  summarise(
    total_sales = sum(amount),
    count = n()
  ) |>
  mutate(
    percentage = (total_sales / sum(total_sales)) * 100
  ) |>
  arrange(desc(total_sales))

# View product summary
product_summary
```

#### Customer Segment Analysis

```{r}
#| label: segment-summary
# Analyze by customer segment
segment_summary <- sales_complete |>
  filter(!is.na(customer_name)) |>
  group_by(segment) |>  # Group by segment
  summarise(
    total_sales = sum(amount),
    customer_count = n_distinct(customer_id),  # Count unique customers
    avg_per_customer = total_sales / customer_count
  )

# View segment summary with formatting
segment_summary |>
  mutate(
    total_sales = dollar(total_sales),
    avg_per_customer = dollar(avg_per_customer)
  )

```

## Key Findings
Data Quality Report

```{r}
#| label: data-quality
# Calculate data quality metrics
total_transactions <- nrow(sales_complete)
valid_transactions <- sum(!is.na(sales_complete$customer_name))
problem_transactions <- sum(is.na(sales_complete$customer_name))

# Create summary
data_quality <- tibble(
  Metric = c("Total Transactions",
            "Valid Transactions",
            "Problem Transactions",
            "Data Quality Rate"),
  Value = c(total_transactions,
            valid_transactions,
            problem_transactions,
            percent(valid_transactions/total_transactions) )
)

data_quality

```


#### Business Insights

Based on the analysis above, complete these insights:

- **Monthly Trend**: Sales (increased) from January to March, with the highest sales in March.
- **Product Performance**: Services generated the most revenue, accounting for 45.8% of total sales.
- **Customer Segments**: The Mid-Market segment had the highest average sale per customer at $33,750.00.
- **Data Quality**: We identified 1 transaction(s) with invalid customer IDs that need investigation.

